### 2010-12-12 Documentation/atropos/atropos.tex
ENH: Updates to bib, tex and eval description --- next commit should reorganize the paper.  TODO: add parameters to theory section.  describe supervised initialization.  describe an application with r-slice data. provide script for this.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@703 fd2da0a4-d968-48df-9cb5-d8c496287cbf

% All other packages
\usepackage{boxedminipage}
\usepackage{graphicx}
\usepackage{amsmath,amsfonts}
\usepackage{babel}
\usepackage{enumerate}
%\usepackage{dsfont}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{sectsty}
\usepackage[compact]{titlesec}
\usepackage[usenames]{color}
\usepackage{ulem}
\usepackage{multirow,booktabs,ctable,array}

@@

Bayes' theorem provides a powerful mechanism for making inductive inferences assuming the availability of quantities defining the relevant conditional probabilities, specifically the likelihood and prior probability terms.  A Bayesian paradigm for brain image segmentation utilizes a user-selected observation model defining the likelihood term and one or more prior probability terms whose product is proportional to the posterior probability.  Whereas the likelihood term has been previously defined parametrically (e.g. a Gaussian model) or non-parametrically (e.g. Parzen windowing of the sample histogram), the prior term, as given in the literature, has been formed inclusively either as MRF-based or template-based.  An image segmentation solution in this context is an assignment of one label to each voxel%
\footnote{
In the classic 3-tissue segmentation case, each voxel in the brain region is assigned a label of `cerebrospinal fluid (csf)', `gray matter (gm)', or `white matter (wm)'. 
}
such that the posterior probability is maximized.  Following
introductory notation, the next sections provide a formal description of these three essential components, viz.
\begin{itemize}
  \item the likelihood or observation model(s),
  \item the prior probability quantities derived from a generalized MRF, template-based prior terms, and a novel distance prior used for label propagation, and 
  \item the optimization framework for maximizing the posterior probability.
\end{itemize}
These key components are common across most segmentation approaches.

\subsection{Notation}

Assume a field, $\mathcal{F}$, whose values are known at discrete locations, i.e. sites, within a regular voxel lattice, $\mathcal{I}$.  Note that $\mathcal{F}$ can be a scalar field in the case of unimodal data (e.g. T1 image only) or a vector field in the case of multimodal data (e.g. T1, T2, and proton density images).
A specific set of observed values, denoted by $\mathbf{y}$, are indexed at $N$ discrete locations in $\mathcal{I}$ by $i \in \{1, 2, \ldots, N\}$.  It is assumed that the complete set of observations is a random field $Y = \{y_1, y_2, \ldots, y_N \}$.    A labeling of $\mathcal{I}$, 
also known as a hard segmentation of $\mathcal{I}$,
assigns to each site in $\mathcal{I}$ one of $K$ labels from the finite set $\mathcal{L} = \{l_1, l_2, \ldots, l_K\}$.  Also considered a random field, we denote this discrete labeling by $X = \{x_1, x_2, \ldots, x_N\}$.

@@

\subsection{Likelihood or Observation Models}

To each of the $K$ labels corresponds a single probabilistic model describing the variation of $\mathcal{F}$ over $\mathcal{I}$.  We denote this set of $K$ likelihood models as $\Phi = \{p_1, p_2, \ldots, p_K\}$.  Using the standard notation, $\mathrm{Pr}(S=s) = p(s)$, $\mathrm{Pr}(S=s|T=t) = p(s|t)$, we can define these voxelwise probabilities, $\mathrm{Pr}_k( Y_i = y_i | X_i = l_k ) = p_k(y_i|l_k)$, in either parametric or non-parametric terms.   Given its simplicity and good performance, in the parametric case, $p_k$ is typically defined as a normalized distribution, i.e.
\begin{align}\label{eq:param}
  p_k\left(y_i|l_k\right) &= G\left(\mu_k;\sigma_k\right) \nonumber \\
                    &= \frac{1}{\sqrt{2\pi \sigma_k^2}}\exp\left( \frac{ -(y_i - \mu_k)^2 }{2\sigma_k^2} \right)
\end{align}
where the parameters $\mu_k$ and $\sigma_k^2$ respectively represent
the mean and variance of the $k^{th}$ model.  In the case of
multi-modal data, where $y_i$ is a vector quantity, a mulitvariate
Gaussian is defined by the mean vector, $\boldsymbol{\mu}_k$, and
covariance matrix, $\Sigma_k$.  An initial estimate for these model
parameters are tabulated during the initialization process,
e.g. K-means or Otsu thresholding.

@@

In order to calculate the probability associated with the entire set of observations, $\mathbf{y}$, we assume independency between voxels which is done primarily for facilitating quantification as opposed to modeling realistic assumptions.  Spatial interdependency between voxels is modeled by the prior probabilities discussed in the next section.  Marginalizing over the set of possible labels, $\mathcal{L}$, leads to the following probabilistic formulation
\begin{align}\label{eq:likelihood}
  p(\mathbf{y}) = \prod_{i=1}^N \left(      
                                                        \sum_{k=1}^K \gamma_k p_k(y_i|l_k)
                                                        \right)
\end{align}
where $\gamma_k$ is the mixing parameter \cite{Ashburner2005}.

\subsection{Prior Probability Models}

By modeling $\mathcal{F}$ via the set of observation models $\Phi$, this so called finite-mixture model could be used to produce a labeling or segmentation \citep[e.g.][]{Wells1996}.  However, as pointed out by \cite{Zhang2001}, the exclusive use of the observation models in modeling the intensity profile produces a less than optimal solution as spatial contextual considerations are completely removed.  This has been remedied by the introduction of a host of  prior probability models including those characterized by their use of MRF theory and template-based information.  For example, in the works of \cite{Leemput1999a} and \cite{Weisenfeld2009}, the original global prior term given in \cite{Wells1996} is replaced by the product of the template-based and the MRF-based prior terms.  In addition to their descriptions below, we discuss our introduction of a third possible prior term in the form of an extension of template-based Euclidean and geodesic distance label propagation.

\subsubsection{Generalized MRF Prior}
In the labeling of a single voxel, an appealing strategy for incorporating spatial coherence into the segmentation solution is to consider the labeling of the voxel neighborhood and favoring labeling configurations in which voxel neighborhoods tended towards homogeneity. This intuition is formally described by MRF theory in which spatial interactions in voxel neighborhoods can be modeled \citep{Li2001}.

In order to apply MRF theory,  we assume the random field introduced earlier, $X$, is an MRF and is, therefore, characterized by a neighborhood system, $\mathcal{N}_i$, on the lattice, $\mathcal{I}$, composed of the neighboring sites of $i$.  This neighborhood system is both noninclusive, i.e. $i \notin \mathcal{N}_i$, and reciprocating, i.e. $i \in \mathcal{N}_j \Leftrightarrow j \in \mathcal{N}_i$.   As an MRF, $X$ also satisfies the positivity and locality conditions, i.e., respectively
\begin{align} \label{eq:mrf}
  p(\mathbf{x}) > 0, \,\, \forall\mathbf{x}
\end{align}
where $\mathbf{x}$ is any particular labeling configuration on $X$ (in
other words, any labeling permutation on $X$ is {\em a priori}
possible).  The MRF locality condition is 
\begin{align}
  p\left(x_i | x_{\mathcal{I}-\{i\}}\right) = p\left(x_i | x_{\mathcal{N}_i}\right),
\end{align}
where $x_{\mathcal{I}-\{i\}}$ is the labeling of the entire image lattice except at site $i$ and  $x_{\mathcal{N}_i}$ is the labeling of $\mathcal{N}_i$.  This locality property enforces solely local considerations based on the neighborhood system in calculating the probability of the particular configuration, $\mathbf{x}$.  Following these two assumptions, the Hammersley--Clifford theorem provides the basis for treating the MRF distribution (cf. Eqn. (\ref{eq:mrf})) as a Gibbs distribution \citep{Besag1974,Geman1984}, i.e.
\begin{align} \label{eq:gibbs}
p(\mathbf{x}) = Z^{-1} \exp\left(-U(\mathbf{x})\right).
\end{align}
$Z$ is a normalization factor known as the {\em partition function}
and $U(\mathbf{x})$ is the {\em energy function} which can take
several forms \citep{Li2001}. In Atropos, as is the case with many
other segmentation algorithms of the same family, we choose
$U(\mathbf{x})$ such that it is only composed of a sum over pairwise
interactions between neighboring sites across the image,%
\footnote{
Using a more expansive definition of  $U(\mathbf{x})$, 
\begin{align}
U(\mathbf{x}) = \sum_{i = 1}^N \left( V_i(x_i) + \beta \sum_{j \in \mathcal{N}_i} V_{ij}( x_i, x_j ) \right), 
\end{align}
would permit casting the other prior terms inside the definition of $U(\mathbf{x})$ in the form of the external field $V_i(x_i)$ but, for clarity purposes, we consider them separately. 
}
i.e.
\begin{align}\label{eq:U}
U(\mathbf{x}) = \beta \sum_{i = 1}^N \sum_{j \in \mathcal{N}_i} V_{ij}( x_i, x_j )
\end{align}
where $V_{ij}$ is typically defined in terms of the Kronecker delta,
$\delta_{ij}$, based on the Ising potential (also known as a Potts model) \citep{Besag1974}
\begin{align}
V_{ij}(x_i, x_j) &= \delta_{ij} \nonumber \\
                          &= \left\{
                          \begin{array}{ll}
                            1 & \text{if } x_i = x_j \\
                            0 & \text{otherwise}
                          \end{array}
                         \right.   
\end{align}
and $\beta$ is a granularity term which weights the contribution of the MRF prior on the segmentation solution.
Since Atropos allows for non-uniform neighborhood systems and systems in which not just the immediate face-connected neighbors are considered, we use the modified function also used in \cite{Noe2001}, which weights the interaction term by the Euclidean distance, $d_{ij}$,  between interacting sites $i$ and $j$ such that 
\begin{align}
  V_{ij} = \frac{\delta_{ij}}{d_{ij}}
\end{align}
so that sites in the neighborhood closer to $i$ are weighted more heavily than sites further away.

\subsubsection{Template-Based Priors}
A number of researchers have used template-based strategies to both ensure spatial coherence and incorporate prior knowledge in obtaining a sutiable segmentation.  A common technique is to select labeled subjects from a population from which a template is constructed \citep[e.g.][which is also available in the ANTs toolkit]{Avants2010}.  Each labeling can then be warped to the template where the synthesis of warped labeled regions produces a prior probability map or prior label map encoding the spatial distribution of labeled anatomy which can be harnessed in joint segmentation/registration or, specifically, Atropos/ANTs strategies involving unlabeled subjects.

We employ the strategy given in \cite{Ashburner2005} in which the
stationary mixing proportions, $\mathrm{Pr}(x_i = l_k) = \gamma_k$,
describing the prior probability that label $l_k$ corresponds to a particular voxel, regardless of intensity, are replaced by the following spatially varying mixing proportions
\begin{align}
\mathrm{Pr}(x_i = l_k) = \frac{\gamma_k t_{ik}}{\sum_{j=1}^K\gamma_j t_{ij}}
\end{align}
where $t_{ik}$ is the prior probability value due to the
template-based prior information at site $i$.  The user may also
choose mixing proportions equal to $\frac{1}{K}$ via the command line
interface to the posterior formulation.
{\textcolor{red}{FIXME --- need explanation of heuristic ...  Also, we
  allow a weighted combination defined by the ComputeLocalPosterior
  function in Atropos --- how do we justify / explain this?   clearly,
it's a balance of the global solution (no local prior) and the local
one with spatial prior ... }}

\subsubsection{Distance Prior for Label Propagation}

@@

\subsection{Optimization}
Given the likelihood models and prior probabilities whose product is proportional to the posterior probability, the Bayesian segmentation solution is the labeling $\hat{\mathbf{x}}$ which maximizes the posterior probability, i.e.
\begin{align}
  \hat{\mathbf{x}} = \argmax_{\mathbf{x}} \left\{p(\mathbf{y}|\mathbf{x})p(\mathbf{x})\right\}.
\end{align}
Similar to its algorithmic predecessors, Atropos employs the EM
framework \citep{Dempster1977}.  After initial estimation of the
likelihood model parameters, EM iterates between calculation of the missing $\hat{\mathbf{x}}$ and subsequent re-estimation of the model parameters by maximizing the expectation of the complete data log-likelihood (cf. Eqn. (\ref{eq:likelihood})).
%\begin{align}
%\log( p(\mathbf{y}) ) = \sum_{i=1}^N \log \left( \sum_{k=1}^K \gamma_k  p(\mathbf{y}|l_k)  \right)
%\end{align}
The expectation maximization procedure is derived in various
publications including \cite{Zhang2001} which yields the optimal mean
and variance (or covariance), but sets the mixing parameter $\gamma_k$
as a constant.  The Atropos implementation estimates $\gamma_k$ at
each iteration as in \cite{Ashburner2005}.
\footnote{
Due to the lack of parameters in the non-parametric approach, it is not technically an EM algorithm (as described in \cite{Wells1996}).  However, the same iterative strategy is applicable and is quite robust in practice as evidenced by the number of researchers employing non-parametric strategies (see the Introduction).
}  
When spatial coherence constraints are included as an MRF prior in Atropos, the optimal segmentation solution becomes intractable.%
\footnote{
Consider $N$ sites each with a possible $K$ labels for a total of $N^K$ possible labeling configurations.  Due to the large $K$ associated with cortical labeling problems, exact optimization is even more problematic than for the traditional 3-tissue scenario.
}
Although many optimization techniques exist \citep[see the
introduction in][for a concise summary of the myriad optimization
possibilities]{Marroquin2002}---each with their characteristic
advantages and disadvantages in terms of computational complexity and
accuracy---Atropos uses the well-known {\em Iterated Conditional
  Modes} (ICM)  \citep{Besag1986} which is greedy, computationally efficient and provides good performance.

@@

As with other classes that comprise ANTs, Atropos uses the Insight Toolkit as a developmental foundation.  
This allows us to take advantage of the mature portions of ITK (e.g. image IO) and ensures the 
integrity of the ancillary processes such as those facilitated by the underlying statistical framework.  Although Atropos is 
publicly distributed with the rest of the ANTs package, we plan to contribute the Atropos segmentation
framework to the Insight Toolkit where it can be vetted and further
improved by other interested researchers. 
{\textcolor{red}{FIXME ---- should we save this for a deliverable in
    some grant?}}

@@

\subsubsection{Initialization}
Atropos was designed to be dimensionally generic in the sense that the same code can be used to handle 2-, 3-, and 
even 4-dimensional image data.  Higher dimensions are possible although we have not yet encountered such 
application-specific need.  Multiple images (assumed to be of the same dimension, size, origin, etc.) can be specified 
on the command line using the `{\ttfamily -a/--intensity-image}' option for each image.  The first image specified on the 
command line is used to initialize the Random, Otsu, or K-means labeling with the latter initialization further
refined by incorporating the additional intensity images, i.e. an initial univariate K-means clustering is determined 
from the first intensity image which, along with the other images, provides the starting multivariate cluster centers for a 
follow-up multivariate K-means labeling.

@@

If the Gaussian likelihood model is chosen, the list sample of intensity values and 
corresponding weights comprised of the posterior probabilities are used to estimate the Gaussian model parameters, i.e.
the mean and variance.  For the non-parametric model, the list sample and posteriors are used
in a Parzen windowing scheme on a weighted histogram to estimate the
observational model  {\textcolor{red} cite Suyash}.

\subsubsection{Prior Probability Models}
Consistent with our previous discussion, we offer both an MRF-based
prior probability for modeling spatial coherence and the possibility
of specifying a set of prior probability maps or a prior label map
with the latter extendable to creating a dense labeling.  To invoke
the MRF `{\ttfamily -m/--mrf}' option, one specifies the smoothing
factor (or the granularity parameter, $\beta$, given in
Eqn. (\ref{eq:U}), and the radius (in voxels) of the neighborhood
system using the vector notation `{\ttfamily 1x1x1}' for a
neighborhood radius of 1 in all 3 dimensions.  Note that this radius
prescribes an $L_\infty$  {\textcolor{red} FIXME should this be $L_2$?} distance and thus includes not just the face-connected neighboring voxels.

>>>>

% All other packages
\usepackage{boxedminipage}
\usepackage{graphicx}
\usepackage{amsmath,amsfonts}
\usepackage{babel,verbatim}
\usepackage{enumerate}
%\usepackage{dsfont}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{sectsty}
\usepackage[compact]{titlesec}
\usepackage[usenames]{color}
\usepackage{ulem}
\usepackage{multirow,booktabs,ctable,array}

@@

Bayes' theorem provides a powerful mechanism for making inductive
inferences assuming the availability of quantities defining the
relevant conditional probabilities, specifically the likelihood and
prior probability terms.  Bayesian paradigms for brain image
segmentation utilize a user-selected observation model defining the
likelihood term and one or more prior probability terms.  The produce
of likelihood(s) and prior(s) is proportional to the posterior
probability.  The likelihood term has been previously defined both 
parametrically (e.g. a Gaussian model) and non-parametrically
(e.g. Parzen windowing of the sample histogram).  The prior term, as
given in the literature, has often been formed either as
MRF-based or template-based.  An image segmentation solution in this
context is an assignment of one label to each voxel%
\footnote{
In the classic 3-tissue segmentation case, each voxel in the brain region is assigned a label of `cerebrospinal fluid (csf)', `gray matter (gm)', or `white matter (wm)'. 
}
such that the posterior probability is maximized.  The next sections
introduce notation and provide a formal description of three essential
components in Bayesian segmentation, viz.
\begin{itemize}
  \item the likelihood or observation model(s),
  \item the prior probability quantities derived from a generalized
    MRF and template-based prior terms, and 
%, and a novel distance prior used for label propagation, and 
  \item the optimization framework for maximizing the posterior probability.
\end{itemize}
These components are common across most EM segmentation approaches.

\subsection{Notation}
Assume a field, $\mathcal{F}$, whose values are known at discrete
locations, i.e. sites, within a regular voxel lattice that makes up an
image domain, $\mathcal{I}$.  Note that $\mathcal{F}$ can be a scalar
field in the case of unimodal data (e.g. T1 image only) or a vector
field in the case of multimodal data (e.g. T1, T2, and proton density
images).  A specific set of observed values, denoted by $\mathbf{y}$,
are indexed at $N$ discrete locations in $\mathcal{I}$ by $i \in \{1,
2, \ldots, N\}$.  This random field, $Y = \{y_1, y_2, \ldots, y_N \}$,
serves as a discrete representation of an observed image's intensities.  A labeling
of this image, also known as a hard segmentation, assigns to each site
in $\mathcal{I}$ one of $K$ labels from the finite set $\mathcal{L} =
\{l_1, l_2, \ldots, l_K\}$.  Also considered a random field, this
discrete labeling is $X = \{x_1, x_2, \ldots, x_N\}$ where each
$x_i \in \mathcal{L}$.   We use $\mathbf{x}$ to denote a specific set of labels in
$\mathcal{I}$ and a valid, though not necessarily optimal, solution to the segmentation problem.

@@

\subsection{Segmentation Objective Function} Atropos optimizes a class
of user-selectable maximization objective functions each of which may
be represented in a generic Bayesian framework, as described by \cite{Sanjay-Gopal1998}.  This framework
requires {\em likelihood models} and {\em prior models} which enter
into Bayes' formula,
\begin{equation}\label{eq:bayes}
p(\mathbf{x}|\mathbf{y})=\underbrace{
p(\mathbf{y}|\mathbf{x})}_{\text{Likelihood(s)}} \underbrace{
p(\mathbf{x})}_{\text{Prior(s)}}\frac{1}{\mathbf{y}}, 
\end{equation} 
where the normalization term, $1/\mathbf{y}$, is a constant that does
not affect the optimization \citep{Sanjay-Gopal1998}.
Given choices for likelihood models and prior probabilities, the Bayesian
segmentation solution is the labeling $\hat{\mathbf{x}}$ which
maximizes the posterior probability, i.e.
\begin{align} \hat{\mathbf{x}} = \argmax_{\mathbf{x}}
\left\{p(\mathbf{y}|\mathbf{x})p(\mathbf{x})\right\}.
\end{align} Similar to its algorithmic predecessors, Atropos employs
the EM framework \citep{Dempster1977} to find maximum likelihood
solutions to this problem.  The following sections detail
Atropos choices for the likelihood(s), prior(s) and EM strategy.

\subsection{Likelihood or Observation Models}
To each of the $K$ labels corresponds a single probabilistic model describing the variation of $\mathcal{F}$ over $\mathcal{I}$.  We denote this set of $K$ likelihood models as $\Phi = \{p_1, p_2, \ldots, p_K\}$.  Using the standard notation, $\mathrm{Pr}(S=s) = p(s)$, $\mathrm{Pr}(S=s|T=t) = p(s|t)$, we can define these voxelwise probabilities, $\mathrm{Pr}_k( Y_i = y_i | X_i = l_k ) = p_k(y_i|l_k)$, in either parametric or non-parametric terms.   Given its simplicity and good performance, in the parametric case, $p_k$ is typically defined as a normal distribution, i.e.
\begin{align}\label{eq:param}
  p_k\left(y_i|l_k\right) &= G\left(\mu_k;\sigma_k\right) \nonumber \\
                    &= \frac{1}{\sqrt{2\pi \sigma_k^2}}\exp\left( \frac{ -(y_i - \mu_k)^2 }{2\sigma_k^2} \right)
\end{align}
where the parameters $\mu_k$ and $\sigma_k^2$ respectively represent
the mean and variance of the $k^{th}$ model.  In the case of
multi-modal data, where $y_i$ is a vector quantity, a mulitvariate
Gaussian is defined by the mean vector, $\boldsymbol{\mu}_k$, and
covariance matrix, $\Sigma_k$.  Multivariate probabilities are derived from the
Mahalanobis distance.  An initial estimate for these model
parameters are tabulated during the initialization process,
e.g. K-means or Otsu thresholding.

@@

Atropos enables two choices for the likelihood formulation.  One is
based on the classical finite mixture model (FMM) and the second is based on
the spatially varying mixture model (SVMM) defined in \citep{Sanjay-Gopal1998}.
In order to calculate the probability associated with the entire set of observations, $\mathbf{y}$, we assume independency between voxels which is done primarily for facilitating quantification as opposed to modeling realistic assumptions.  Spatial interdependency between voxels is modeled by the prior probabilities discussed in the next section.  Marginalizing over the set of possible labels, $\mathcal{L}$, leads to the following probabilistic formulation
\begin{align}\label{eq:likelihood}
  p(\mathbf{y}|\mathbf{x}) = \prod_{i=1}^N \left(      
                                                        \sum_{k=1}^K \gamma_k p_k(y_i|l_k)
                                                        \right)
\end{align}
where $\gamma_k$ is the mixing parameter \citep{Ashburner2005}.

\subsection{Prior Probability Models}
By modeling $\mathcal{F}$ via the set of observation models $\Phi$,
this so called finite-mixture model could be used to produce a
labeling or segmentation \citep[e.g.][]{Wells1996}.  However, as
pointed out by \cite{Zhang2001}, the exclusive use of the observation
models in modeling the intensity profile produces a less than optimal
solution because spatial contextual considerations are ignored.  This
has been remedied by the introduction of a host of prior probability
models including those characterized by use of MRF theory and
template-based information.  For example, in the works of
\cite{Leemput1999a} and \cite{Weisenfeld2009}, the original global
prior term given in \cite{Wells1996} is replaced by the product of the
template-based and the MRF-based prior terms.  In addition to their
descriptions below, we discuss our introduction of a third possible
prior/objective combination in the form of a (sparse) prior labeling which fixes specific points
of the segmentation and uses EM to propagate this information
elsewhere in the image.  
% extension of template-based Euclidean and geodesic distance label propagation.

\subsubsection{Generalized MRF Prior}
One may incorporate spatial coherence into the segmentation by
favoring labeling configurations in which voxel neighborhoods tend
towards homogeneity. This intuition is formally described by MRF
theory in which spatial interactions in voxel neighborhoods can be
modeled \citep{Li2001}.

We assume the random field introduced earlier, $X$, is an MRF characterized by a neighborhood system, $\mathcal{N}_i$, on the lattice, $\mathcal{I}$, composed of the neighboring sites of $i$.  This neighborhood system is both noninclusive, i.e. $i \notin \mathcal{N}_i$, and reciprocating, i.e. $i \in \mathcal{N}_j \Leftrightarrow j \in \mathcal{N}_i$.   As an MRF, $X$ also satisfies the positivity and locality conditions, i.e., 
 \begin{align} \label{eq:mrf}
  p(\mathbf{x}) > 0, \,\, \forall\mathbf{x} 
 \end{align}
and where $\mathbf{x}$ is any particular labeling configuration on $X$ (in
other words, any labeling permutation on $X$ is {\em a priori}
possible).  The MRF locality condition is then, 
\begin{align}
  p\left(x_i | x_{\mathcal{I}-\{i\}}\right) = p\left(x_i | x_{\mathcal{N}_i}\right),
\end{align}
where $x_{\mathcal{I}-\{i\}}$ is the labeling of the entire image lattice except at site $i$ and  $x_{\mathcal{N}_i}$ is the labeling of $\mathcal{N}_i$.  This locality property enforces solely local considerations based on the neighborhood system in calculating the probability of the particular configuration, $\mathbf{x}$.  Following these two assumptions, the Hammersley--Clifford theorem provides the basis for treating the MRF distribution (cf. Eqn. (\ref{eq:mrf})) as a Gibbs distribution \citep{Besag1974,Geman1984}, i.e.
\begin{align} \label{eq:gibbs}
p(\mathbf{x}) = Z^{-1} \exp\left(-U(\mathbf{x})\right).
\end{align}
$Z$ is a normalization factor known as the {\em partition function}
and $U(\mathbf{x})$ is the {\em energy function} which can take
several forms \citep{Li2001}. In Atropos, as is the case with many
other segmentation algorithms of the same family, we choose
$U(\mathbf{x})$ such that it is only composed of a sum over pairwise
interactions between neighboring sites across the image,%
\footnote{
Using a more expansive definition of  $U(\mathbf{x})$, 
\begin{align}
U(\mathbf{x}) = \sum_{i = 1}^N \left( V_i(x_i) + \beta \sum_{j \in \mathcal{N}_i} V_{ij}( x_i, x_j ) \right), 
\end{align}
would permit casting the other prior terms inside the definition of $U(\mathbf{x})$ in the form of the external field $V_i(x_i)$ but, for clarity purposes, we consider them separately. 
}
i.e.
\begin{align}\label{eq:U}
U(\mathbf{x}) = \beta \sum_{i = 1}^N \sum_{j \in \mathcal{N}_i} V_{ij}( x_i, x_j )
\end{align}
where $V_{ij}$ is typically defined in terms of the Kronecker delta,
$\delta_{ij}$, based on the classical Ising potential (also known as a Potts model) \citep{Besag1974}
\begin{align}
V_{ij}(x_i, x_j) &= \delta_{ij} \nonumber \\
                          &= \left\{
                          \begin{array}{ll}
                            1 & \text{if } x_i = x_j \\
                            0 & \text{otherwise}
                          \end{array}
                         \right.   
\end{align}
and $\beta$ is a granularity term which weights the contribution of the MRF prior on the segmentation solution.
Since Atropos allows for non-uniform neighborhood systems and systems in which not just the immediate face-connected neighbors are considered, we use the modified function also used in \cite{Noe2001}, which weights the interaction term by the Euclidean distance, $d_{ij}$,  between interacting sites $i$ and $j$ such that 
\begin{align}
  V_{ij} = \frac{\delta_{ij}}{d_{ij}}
\end{align}
so that sites in the neighborhood closer to $i$ are weighted more heavily than sites further away.

\subsubsection{Template-Based Priors}
A number of researchers have used template-based strategies to both ensure spatial coherence and incorporate prior knowledge in segmentation.  A common technique is to select labeled subjects from a population from which a template is constructed \citep[e.g.][which is also available in the ANTs toolkit]{Avants2010}.  Each labeling can then be warped to the template where the synthesis of warped labeled regions produces a prior probability map or prior label map encoding the spatial distribution of labeled anatomy which can be harnessed in joint segmentation/registration or, specifically, Atropos/ANTs strategies involving unlabeled subjects.

We employ the strategy given in \cite{Ashburner2005} in which the
stationary mixing proportions, $\mathrm{Pr}(x_i = l_k) = \gamma_k$,
FIXME DEFINE THE GAMMA --- SHOULD USE SVMM FROM HEBERT ?
describing the prior probability that label $l_k$ corresponds to a particular voxel, regardless of intensity, are replaced by the following spatially varying mixing proportions
\begin{align}
\mathrm{Pr}(x_i = l_k) = \frac{\gamma_k t_{ik}}{\sum_{j=1}^K\gamma_j t_{ij}}
\end{align}
where $t_{ik}$ is the prior probability value due to the
template-based prior information at site $i$.  The user may also
choose mixing proportions equal to $\frac{1}{K}$ via the command line
interface to the posterior formulation.

% {\textcolor{red}{FIXME --- need explanation of heuristic ...  Also, we

%  allow a weighted combination defined by the ComputeLocalPosterior

%  function in Atropos --- how do we justify / explain this?   clearly,

% it's a balance of the global solution (no local prior) and the local

% one with spatial prior ... }}

\subsubsection{Supervised Initialization to the EM Algorithm}
PriorLabelImage

\begin{comment}
{
\subsubsection{Distance Prior for Label Propagation}

@@

}
\end{comment}

\subsection{Optimization}
Atropos uses expectation maximization to find a locally optimal
solution for the user-selected version of the Bayesian segmentation
problem (cf. Eqn. (\ref{eq:bayes})).
After initial estimation of the
likelihood model parameters, EM iterates between calculation of the
missing $\hat{\mathbf{x}}$ and subsequent re-estimation of the model
parameters by maximizing the expectation of the complete data
log-likelihood (cf. Eqn. (\ref{eq:likelihood})).  
%\begin{align}
%\log( p(\mathbf{y}) ) = \sum_{i=1}^N \log \left( \sum_{k=1}^K \gamma_k  p(\mathbf{y}|l_k)  \right)
%\end{align}
The expectation maximization procedure is derived in various
publications including \cite{Zhang2001} which yields the optimal mean
and variance (or covariance), but sets the mixing parameter $\gamma_k$
as a constant.  The Atropos implementation estimates $\gamma_k$ at
each iteration as in \cite{Ashburner2005}.
\footnote{
Due to the lack of parameters in the non-parametric approach, it is not technically an EM algorithm (as described in \cite{Wells1996}).  However, the same iterative strategy is applicable and is quite robust in practice as evidenced by the number of researchers employing non-parametric strategies (see the Introduction).
}  
When spatial coherence constraints are included as an MRF prior in Atropos, the optimal segmentation solution becomes intractable.%
\footnote{
Consider $N$ sites each with a possible $K$ labels for a total of $N^K$ possible labeling configurations.  Due to the large $K$ associated with $K>>1$ problems, exact optimization is even more problematic than for the traditional 3-tissue scenario.
}
Although many optimization techniques exist \citep[see the
introduction in][for a concise summary of the myriad optimization
possibilities]{Marroquin2002}---each with their characteristic
advantages and disadvantages in terms of computational complexity and
accuracy---Atropos uses the well-known {\em Iterated Conditional
  Modes} (ICM)  \citep{Besag1986} which is greedy, computationally
efficient and provides good performance.  The EM algorithm employed in
Atropos may therefore be written
\begin{eqnarray}
X \\ 
Y \\
Z
\end{eqnarray}
The solution will be a local maxima.

@@

As with other classes that comprise ANTs, Atropos uses the Insight Toolkit as a developmental foundation.  
This allows us to take advantage of the mature portions of ITK (e.g. image IO) and ensures the 
integrity of the ancillary processes such as those facilitated by the underlying statistical framework.  Although Atropos is 
publicly distributed with the rest of the ANTs package, we plan to contribute the Atropos segmentation
framework to the Insight Toolkit where it can be vetted and further
improved by other interested researchers.

@@

\subsubsection{Initialization}
Atropos was designed to be dimensionally generic in the sense that the same code can be used to handle 2-, 3-, and 
even 4-dimensional image data.  Higher dimensions are possible although we have not yet encountered such 
application-specific need.  Multiple images (assumed to be of the same
dimension, size, origin, etc.), which enable multivariate likelihoods, can be specified 
on the command line using the `{\ttfamily -a/--intensity-image}' option for each image.  The first image specified on the 
command line is used to initialize the Random, Otsu, or K-means labeling with the latter initialization further
refined by incorporating the additional intensity images, i.e. an initial univariate K-means clustering is determined 
from the first intensity image which, along with the other images, provides the starting multivariate cluster centers for a 
follow-up multivariate K-means labeling.

@@

If the Gaussian likelihood model is chosen, the list sample of intensity values and 
corresponding weights comprised of the posterior probabilities are used to estimate the Gaussian model parameters, i.e.
the mean and variance.  For the non-parametric model, the list sample and posteriors are used
in a Parzen windowing scheme on a weighted histogram to estimate the
observational model  \citep{Awate2006}.

\subsubsection{Prior Probability Models}
Consistent with our previous discussion, we offer both an MRF-based
prior probability for modeling spatial coherence and the possibility
of specifying a set of prior probability maps or a prior label map
with the latter extendable to creating a dense labeling.  To invoke
the MRF `{\ttfamily -m/--mrf}' option, one specifies the smoothing
factor (or the granularity parameter, $\beta$, given in
Eqn. (\ref{eq:U}), and the radius (in voxels) of the neighborhood
system using the vector notation `{\ttfamily 1x1x1}' for a
neighborhood radius of 1 in all 3 dimensions.  This radius is defined
such that voxels including but not limited to those that are face
connected will influence the MRF.
